### 2005-12-13 inst/paper/paper2.tex
Updated paper 2.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@15362 bc3139a8-67e5-0310-9ffc-ced21a209358

\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}

@@

\section{Introduction}\label{Sec:Intro}

@@

There are many ways to visualize multivariate data where more than
two or three dimensions are needed to view the data.  Several of these ideas
were discussed in a previous paper, titled ``Creating Linked, Interactive
Plots to Explore Multivariate Data'', so the focus in this paper will be one
particular method of visualizing multivariate data that is particularly useful
for exploratory data analysis.  This paper will discuss creating linked,
interactive views of linked data sets.

First brief definitions of linked and interactive views will be given.  Linked
plots mean that if a point on one plot changes its appearance, such as it
becomes colored red, then the corresponding point on a second plot will also
change its appearance to the color red.  For points to be corresponding, the
data displayed in the points must come from the same entity.  As an example,
suppose that the data consisted of each of the 50 states' murder rate, assault
rate, rape rate and percent of population living in urban areas.  Then if
there were two scatterplots, one with murder rate versus assault rate and the
other with rape rate versus percent urban population, the points on each plot
that referred to the state of Montana would be corresponding points.

@@

Creating interactive applications requires three
things: some user action or input, which will be referred to as an
event; a response to that action, which will be executed by a callback
function; and a method that connects the event to the callback
function, which will be referred to as the signal handler.  These
three steps are ordered as follows: an event occurs, the event causes a
signal to be emitted that is caught by the signal handler, and then
the signal handler calls the callback function, which results in the
response to the user action.  A flowchart of these steps is shown
below in Figure \ref{Fig:Event}.  In the example of clicking on a
hyperlink mentioned above, the event is the user clicking on the
hyperlink and the callback function causes a new web page to open.
The signal handler, which is the method that connects the event to the
callback function, is not noticed by the user.

\begin{figure}[!h]
  \begin{center}    
    \scalebox{0.8}{\includegraphics{Event.jpg}}
    \caption{ Flow Chart of the Response to an Event. }
    \label{Fig:Event}
  \end{center}
\end{figure}

This method of creating linked, interactive views of data is flexible for
exploring multivariate data because many different views can be created, such
as scatterplots, spreadsheets, and heatmaps, and then these views can be
linked.  Thus, users can decide which views best visually represent their
data.  Also, since the views are interactive, users can change the views while
looking at them to make the views more informative about the underlying data
structure.  This flexibility in linked, interactive views gives users a
powerful tool for visually exploring their data.

@@

Focusing on linked, interactive views as a powerful visualization tool, the
next step is to determine how to design a software package that can implement
this method.  A popular design for creating linked views of data is the
model-view-controller (MVC) paradigm.  This design consists of three types
of objects: the controller, which defines what actions occur in response to
user input; the view, which consists of displays of the data; and the model,
which manages the data.  This paradigm is powerful because it decouples the
views from the model.  Thus, only one copy of the data needs to be stored and
then all views refer to this data set.  With all views based on the same data,
any changes to the data will be propagated to all views, which will give the
appearance of linked views.

To link the views to this one data set there is a subscribe/notify procedure
between the views and the model.  The views subscribe to a particular model
and the model must notify the views when a change occurs so that the views
will be updated.  This design allows multiple views of the model that are
linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{ExMVC.jpg}}
    \caption{ A Model with its Views. }
    \label{Fig:ExMVC}
  \end{center}
\end{figure}

In the ``Creating Linked, Interactive Plots to Explore Multivariate Data''
paper, the creation of an R package called \Rpackage{iSPlot} was discussed.
The \Rpackage{iSPlot} package implemented the MVC design to satisfy three
goals: to create views that are linked, to create views that are interactive,
and to create a design that is extensible.  This package currently allows
users to generate linked, interactive scatterplots and spreadsheets of
two-dimensional data, such as data frames and matrices.

@@

The MVC design assumes that there is one data set that all views will
reference.  But what if the data consisted of more than one data set?  For
example, consider the data stored in a relational database.  Here there are
often multiple tables of data, where rows in different tables are linked
through a unique identifier called a key.  Although these tables of data
could be combined into one large table, there are several reasons not to do
this.  Keeping the data in separate tables reduces redundancy, particularly
when the rows are related by a one-to-many or a many-to-many relationship.
Also separate data tables lead to more maintainable data.  By storing only one
copy of each piece of information (in a row in one table), it is easier to
ensure that the data are correctly updated when a change occurs.  All other
tables that are connected with this information can then just reference this
row that has been updated.

Consider Figure \ref{Fig:DBTab} shown below.  Here the database consists of
three tables, Students, Classes, and StudentClasses.  The Students table has
four columns, StudentID (the key), FirstName, LastName, and Year.  The Classes
table has three columns, ClassID (the key), ClassName, and Department.  The
StudentClasses table has two columns, StudentID and ClassID.  This last table
relates which classes a student is taking and which students are in
each class.  In this example, the students can be taking more than one class
and each class can have more than one student.  So the relationship between
classes and students is many-to-many.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{databaseTables.jpg}}
    \caption{ Example of Relational Database Tables. }
    \label{Fig:DBTab}
  \end{center}
\end{figure}

Clearly, all of this data could be combined into one large
table, called CombinedStudentClass, as shown in Figure \ref{Fig:OneDBTab}.
The CombinedStudentClass table would have the following columns: StudentID,
FirstName, LastName, Year, ClassID, ClassName, and Department.  Now if a
student's last name changed, all instances of this student would need to be
changed in the new large table.  For example, if the student was taking four
classes, then this student would have four rows in the CombinedStudentClass
table and each row would need to change the value for the student's last name.
If instead only one record was kept per student, as shown in Figure
\ref{Fig:DBTab} above, then this process would be less error prone and would
require less memory to store the data. This example shows that there are
advantages to storing the data in separate structures.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{oneDatabaseTable.jpg}}
    \caption{ Example of a Combined Table. }
    \label{Fig:OneDBTab}
  \end{center}
\end{figure}

In a relational database all of the data are stored in tables, but consider the
situation where there are multiple data sets that are of different data
structures.  For example, suppose that you have experimental microarray data,
which looks at gene expression levels.  This experimental data can be
stored as a two dimensional data structure, such as a matrix, where rows are
genes and columns are samples.  Linked to this
experimental data is meta data that gives information about the genes studied
in the microarray experiment.  This meta data may be a gene ontology
graph that gives the molecular functions of these genes.  These two data sets
(the experimental data and the meta data) are linked through the genes, but
they are different data structures.  The experimental data is a matrix and
the meta data is a graph.

Now for the original MVC design, where one data set is expected, it is not
clear how to combine these two data sets into one data set.  The meta data
graph could be stored as a matrix, but a graph is a data structure that can be
well represented as an object with slots for nodes and edges, especially if
the graph is sparse.  As with the relational database tables, it seems
best in terms of data storage, alteration, and retrieval to keep the two data
sets separate.  Also, keeping the two data sets separate does not force a data
structure on the data that may not fit well.

Currently, linked data sets are a common occurrence that can not be easily
visualized simultaneously.  Any time there is meta data linked to experimental
or study data there is a multiple data set situation where visually exploring
the data would be helpful to understand relationships between variables.  In
this paper, the design of software to create linked views based on multiple
linked data sets will be discussed.  The goals of this software will include
creating linked views of linked data sets, creating interactive
views where the response to an event can be changed, and creating
an extensible design that will allow future users to make additions.

@@

As discussed in Section \ref{Sec:Intro}, the MVC design expects one data set
with all views referencing this one data set.  However, several examples were
shown where multiple data sets are linked.  Since linking between multiple
data sets would be useful to visualize, the next step will be to determine how
the MVC design can be revised to account for multiple data sets.

@@

One possibility for dealing with multiple data sets within the MVC design is to
combine the multiple data sets into one large data set.  With this new
combined data, the original MVC design will be appropriate for creating linked
views.  Thus, the strength of combining the data sets into one large data set
is that the MVC paradigm will already be appropriately set up to deal with the
one large data set.

Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
multiple data sets into one large data set is not always appropriate or even
possible.  Even if the multiple data sets have the same data structure,
combining the data sets often results in redundant data and makes updating the
data more complicated.  In the situation where the data sets are not of the
same data structure, it may not be possible to combine the data sets into one
data structure that retains all of the original information.  Thus, another
way of linking views based on multiple data sets needs to be designed.

@@

Because of the difficulty of combining multiple data sets, the MVC paradigm
will need to be altered to fit the data.  It has been mentioned that the MVC
design is intended to deal with one data set.  Now that we have several data
sets, why not have several MVC objects?  One MVC object to handle each data
set.  Thus, within each MVC object, one could create linked views of that
particular data set, which is the same as the original MVC design.  Then
between MVC objects a messaging system would need to be designed so that a
change in one data set would be sent to the other linked data sets.

@@

\normalsize
As an example, suppose that one has two data sets: one with information at the
state level and one with information at the city level.  These tables are
shown in Table \ref{Tab:CityState}.  The State data includes Life Expectancy
(in years from 1970) and Income (per capita from 1974) data
while the City data includes Precipitation (in inches from 1975) and Average
High Temperature in January (in Fahrenheit from 2005)
data.  These two data sets are linked by the CityState table that links cities
with the states they reside in.  Now for these two data sets there would be
two MVC objects.  One MVC object for the State data and its views and one MVC
object for the City data and its views.

@@

Two parts of this design will need some special thought: creating an
appropriate messaging system between MVC objects and deciding how the MVC
objects will be related.  First, there are several questions that need to be
considered when creating the message system.  What information should be
passed between MVC objects in the messages?  Should there be a system to stop
messages from propagating between MVC objects?  How will the messages be
created and handled?  Second, there are also several questions with regard to
how the MVC objects will be related.  Should the MVC objects be related like
the tables in a relational database where the relationships are determined by
linked keys in different tables?  Should the MVC objects be linked through some
type of hierarchy, like a parent-child relationship where one MVC creates
another MVC and thus, the original MVC is the parent?  These decisions will be
based on the data that needs to be visualized and will be discussed in the
following sections as the design for multiple MVC objects is explained.

@@

Implementing a design to create linked, interactive views of multiple linked
data sets will be done in the R language in two packages, called
\Rpackage{MVCClass} and \Rpackage{iSNetwork}.  The \Rpackage{MVCClass} package
will implement the design of the MVC classes and methods.  The
\Rpackage{MVCClass} package will define the individual model, view, and
controller classes as well as the \Robject{MVC} class that binds the
individual model, view, and controller classes together into one unit.  The
package will also define the message classes that will allow the components to
communicate with each other.  The goal of separating these class definitions
and methods into one package is that they will be available for other R
packages that implement linked views of data, such as the \Rpackage{iSPlot}
and the \Rpackage{iSNetwork} packages.

With the class definitions created in the \Rpackage{MVCClass} package, the
\Rpackage{iSNetwork} package will implement the graphical user interface
(GUI) so that users can load data, create views, and interact with the views
through a user interface with menus.  To create the GUI, \Rpackage{iSNetwork}
will use the R packages, \Rpackage{RGtk} and \Rpackage{gtkDevice}.
\Rpackage{RGtk} is an R package that allows one to interact with Gtk functions
from the R interface.  Gtk is an open-source X Window toolkit for creating
user interfaces.  The \Rpackage{gtkDevice} package creates a Gtk device by
turning a Gtk drawing area into a device that acts like an R device, but still
responds to events so that the view can be interactive.  For more information
about these packages, please see the paper, ``Creating Linked, Interactive
Plots to Explore Multivariate Data''.

The next subsections will explain the class definitions and methods that
pertain to a single model-view-controller object.  Thus, the model classes,
the view classes, the controller, the \Robject{MVC} class (which binds the
model, view and controller classes together), and the message classes that pass
information between the three components in one MVC object will all be
discussed in the following subsections.  All of these definitions are
implemented in the \Rpackage{MVCClass} package.  Message class definitions
that pertain to communication between the MVC objects will be given in
Section \ref{Sec:MultMVC}.

@@

The model class is responsible for storing and updating the data.  These two
functions are reflected in Figure \ref{Fig:Model}, which shows the inheritance
structure for the model classes.  Here the virtual class,
\Robject{gModel}, has the slots modelData, linkData, virtualData, and
modelName.  These four slots are common information that all models will
need.  The modelData slot is the data set for this model, the linkData slot is
a list of two functions, \Rfunction{linkToParent} and \Rfunction{linkToChild},
which link this model to its parent and child models, respectively (see Section
\ref{Ssec:MultLink}), the virtualData slot is data pertaining to the views
that needs to be stored with the model so that it can be shown in all views of
this model, and the modelName slot is the name of the model.

@@

For specific model classes, there are \Robject{dfModel}, \Robject{graphModel},
and \Robject{exprModel}.  The \Robject{dfModel} class represents a model where
the modelData has class data frame (or matrix), the \Robject{graphModel} class
represents a model where the modelData has class graph, and the
\Robject{exprModel} class represents a model where the modelData has class
exprSet.

As mentioned previously, the model classes are also responsible for updating
the data and thus, you will notice that all three model classes have an
\Rfunction{updateModel} method.  This method is called whenever the data needs
to be updated in response to a user interacting with a view.  Thus, the
\Rfunction{updateModel} method is actually updating the virtualData slot
because it is this slot in the model classes that ensures that all views are
appropriately visually linked.

Although only three model classes are currently implemented, future users are
welcome to make additions to the model classes for new data structures they
would like to represent.  For example, a user may want to add a new model
class called \Robject{tsModel} to represent time series data.

Note that the model classes must also notify views when the data has changed
so that the views can be updated.  The communication between a model and its
views will be discussed in Section \ref{Ssec:OneMess}.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{ViewClass.jpg}}
    \caption{ Inheritance for View Objects. }
    \label{Fig:View}
  \end{center}
\end{figure}

>>>>

\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rslot}[1]{\textsl{#1}}

@@

\bibliographystyle{amsplain}
\bibliography{paper2}

\begin{titlepage}
\begin{center}
\Large
\vspace*{1in}
\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
\vspace{1in}
\textbf{Dana-Farber Cancer Institute}\\ 
\textbf{December 9, 2005} \\
\vspace{1in} 
\textbf{Elizabeth Whalen} \\
\end{center}
\end{titlepage}
\normalsize
\newpage

%\renewcommand{\baselinestretch}{1.5}
\section{Introduction}\label{Sec:Intro}

@@

There are many ways to visualize multivariate data where more than
two or three dimensions are needed to view the data.  Several of these ideas
were discussed in a previous paper, titled ``Creating Linked, Interactive
Views to Explore Multivariate Data'' \cite{EW05}, so the focus in this paper
will be one specific method of visualizing multivariate data that is
particularly useful for exploratory data analysis.  This paper will discuss
creating linked, interactive views of linked data sets.

First brief definitions of linked views, interactive views and linked data
sets will be given.  Linked views mean that if a component that represents
data (such as a point on a plot or a row in a spreadsheet) changes its
appearance on one view, then the corresponding component on a second view will
also change its appearance.  For components on different views to be
corresponding, the data displayed in those components must come from the same
entity in a model.  As an example, suppose that the model consisted of four
variables for the fifty states' murder rate, assault rate, rape rate, and
percent of population living in urban areas.  Now the model could be
represented as a matrix of four columns by fifty rows with four columns for
the four variables and fifty rows for the fifty states.  In this model, the
entities would be the states.  If two scatterplots were created as views of
this model, where the first plot displayed murder rate versus assault rate and
the second plot displayed rape rate versus percent urban population, then the
points on each plot that referred to the state of Montana would be
corresponding components.

@@

One particular model for creating interactive applications involves three
components: some user action or input, which will be referred to as an event;
a response to that action, which will be executed by a function that is called
a callback function; and a method that connects the event to the callback
function, which will be referred to as the signal handler.  The events can
range from the user moving the cursor over a plot to pressing a button on a
graphical user interface (GUI) to pressing certain keys.  Notice that the
events almost always occur over some type of interface, though some events,
such as key press events may occur without an interface.  The callback
function is a unique function because its signature is determined by the event
that the callback function is linked to.  The parameters of the callback
function may include information about the event, such as the location where
the event occurred, and information about the object that just caused the
event, such as the button on the GUI.  The signal handler is the piece that
must look to see if an event has occurred and then when the event has
occurred, the signal handler calls the appropriate callback function.  The
strengths of this model of interactivity are that programmers can decide which
events to notice by deciding which signal handlers to define, programmers
can change the response to an event by changing the callback function that is
linked to an event, and the system's event loop can continue working on other
things until a signal is emitted because an event occurred.

These three pieces in this model of interactivity are ordered as follows:
an event occurs, the event causes a signal to be emitted that is caught by the
signal handler, and then the signal handler calls the callback function, which
results in the response to the user action.  A flowchart of these steps is
shown below in Figure \ref{Fig:Event}.  In the example of clicking on a
hyperlink mentioned above, the event is the user clicking on the
hyperlink and the callback function causes a new web page to open.
The signal handler, which is the method that connects the event to the
callback function, is not noticed by the user, but it is handled by the
system's event loop when a signal is emitted.

\begin{figure}[!h]
  \begin{center}    
    \scalebox{0.6}{\includegraphics{Event.jpg}}
    \caption{ Flow Chart of the Response to an Event. }
    \label{Fig:Event}
  \end{center}
\end{figure}

Linked data sets mean that the data have more than one conceptual grouping,
but there is a relationship between these different conceptual groupings.
Linked data sets are an idea that will be familiar to users of relational
databases.  Data in a relational database are typically stored in many
linked tables with rows in different tables linked to each other through a
unique identifier called a key.  In this example, each table would be
considered a data set and they are linked through the keys.  Another situation
where linked data sets occur is when there is experimental or study data that
are linked to meta-data.  An example of this situation is given in Section
\ref{Ssec:Limit} where the experimental data are microarray data, which looks
at gene expression levels, and the meta-data are a Gene Ontology graph that
gives the molecular functions of the genes studied in the microarray
experiment.

Creating linked, interactive views of data is flexible for exploring
multivariate data because many different views can be created, such as
scatterplots, spreadsheets, and heatmaps, and these views are linked because
they are based on the same underlying model.  Thus, users can decide which
views best visually represent their data.  Also, since the views are
interactive, users can change the views while looking at them to make the
views more informative about the underlying data structure.  This flexibility
in linked, interactive views gives users a powerful tool for visually
exploring their data.

In Section \ref{Ssec:Design}, the design of linked, interactive views based on
the model-view-controller (MVC) paradigm will be discussed.  However, the
limitation with this design is that it expects one data set, which limits the
type of data we can visualize.  There are many
situations where multiple data sets occur, particularly any time there are
experimental or study data linked to meta-data.  Being able to visualize data
from multiple data sets would be helpful to understand the relationships
between variables.  Thus, after discussing the limitations of the MVC design
in Section \ref{Ssec:Limit}, an extension of the MVC design will be discussed
in Section \ref{Sec:Extend}.  This extension will allow the creation of linked,
interactive views based on linked data sets.  The software that is discussed
in this paper will have three goals: to create linked views of linked data
sets, to create interactive views where the response to an event can be
changed, and to create an extensible design that will allow future users to
make additions.

As a motivating example of linked views based on linked data sets, consider
Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
When a user initially interacts with the scatterplot view of data set 2, the
response to this interaction is to tell data set 2 to update itself (this is
indicated on the figure as a circled 1).  Then in step 2, data set 2 tells the
scatterplot view that is should update itself in response to the change in the
data.  Next, in step 3, because data set 2 has changed and because it is
linked to data set 1, data set 2 sends information to data set 1 about the
change.  In step 4, the information from data set 2 is converted into
information that data set 1 can understand.  In step 5, this information is
received by data set 1 and data set 1 updates its data.  Finally, in step 6,
data set 1 tells its view to update itself.  Thus, in this example, the views
are synchronized based on the underlying linked data sets.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{firstMessagePassing.jpg}}
    \caption{ Linked Views of Linked Data Sets. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

Focusing on linked, interactive views as a powerful visualization tool, the
next step is to determine how to design a software package that can implement
this method.  A popular design for creating linked views of data is the
model-view-controller (MVC) paradigm.  This design consists of three types
of objects: the controller, which defines what actions occur in response to
user input; the view, which consists of displays of the data; and the model,
which manages the data.  This paradigm is powerful because it decouples the
views from the model.  Thus, only one copy of the data needs to be stored and
then all views refer to this data set.  With all views based on the same data,
any changes to the data can be propagated to all views, which will create
linked views.

To link the views to this one data set there is a subscribe/notify procedure
between the views and the model.  The views subscribe to a particular model
and the model must notify the views when a change occurs so that the views
will be updated.  This design allows multiple views of the model that are
linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}
\cite{DesignPatterns}.  Each view depends only on visually representing the
model and thus, it is not necessary for a view to be aware of other views.  By
keeping each view ignorant of other views, other views can added and removed
without affecting a different view.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{ExMVC.jpg}}
    \caption{ A Model with its Views. }
    \label{Fig:ExMVC}
  \end{center}
\end{figure}

Notice that although the components (model, view, and controller) in the MVC
design are independent of one other, there still needs to be communication
between them so that information can be shared.  Examples of this
communication between components include the model notifying views that the
data have changed, a view asking the controller how it should respond to user
interaction, and a view telling the model that the data should change in
response to user interaction with the view.  This message passing between
components in the MVC paradigm will be discussed in detail in Section
\ref{Ssec:OneMess}.

The ``Creating Linked, Interactive Views to Explore Multivariate Data''
\cite{EW05} paper discussed the creation of R software to fulfill three goals:
to create views that are linked, to create views that are interactive, and to
create a design that is extensible.  To satisfy the overall goals of the
software, two R packages, called \Rpackage{MVCClass} and \Rpackage{iSPlot},
were created.  The \Rpackage{MVCClass} package implemented the class and
generic function definitions needed for the MVC design and the
\Rpackage{iSPlot} package created the front end to let the user create linked,
interactive views of two-dimensional data, such as data frames and matrices,
through either a GUI or through command line functions.  The reason for
separating the software into two packages was that it allowed the class
definitions in the \Rpackage{MVCClass} package to be used by any other R
package that would like to create linked views.

@@

The MVC design assumes that there is one data set that all views will
reference.  But what if the data consisted of more than one conceptual
grouping?  For example, consider data stored in a relational database, where
there are multiple tables of data and rows in different tables are linked
through a unique identifier called a key.  Although these tables of data could
be combined into one large table, there are several reasons not to do this.
Keeping the data in separate tables reduces redundancy, particularly when the
rows are related by a one-to-many or a many-to-many relationship.  Also
separate data tables generally lead to more maintainable data.  By storing
only one copy of each piece of information (in a row in one table), it is
easier to ensure that the data are correctly updated when a change occurs.
All other tables that are connected with this information can then just
reference this row that has been updated.

Consider Figure \ref{Fig:DBTab} shown below.  Here the database consists of
three tables, State, City and StateCity.  The State table has three columns,
State (the key), Life Expectancy, and Income.  The City table has three
columns, City (the key), Precipitation, and Temperature.  The StateCity table
has two columns, State and City.  This last table relates which cities are in
which states.  In this example, each city can be in only one state, but each
state can have more than one city so there is a one-to-many relationship
between states and cities.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newDatabaseTables.jpg}}
    \caption{ Example of Relational Database Tables. }
    \label{Fig:DBTab}
  \end{center}
\end{figure}

Clearly, this information could be combined into one large table, called
CombinedStateCity, as shown in Figure \ref{Fig:OneDBTab}.  The
CombinedStateCity table would have the following columns: State, Life
Expectancy, Income, City, Precipitation, and Temperature.  Now if a state's
life expectancy changed, all instances of this state would need to be changed
in the new large table.  For example, if the state had the information for
three cities recorded, then the state would have three rows in the
CombinedStateCity table and each row would need to change the value for the
state's life expectancy.  If instead only one record was kept per state, as
shown in Figure \ref{Fig:DBTab} above, then this process would be less error
prone and would require less memory to store the data.  This example shows
that there are advantages to storing the data in separate structures.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newOneDatabaseTable.jpg}}
    \caption{ Example of a Combined Table. }
    \label{Fig:OneDBTab}
  \end{center}
\end{figure}

In a relational database all of the data are stored in tables, but consider the
situation where there are multiple data sets that are of different data
structures.  For example, suppose that you have experimental microarray data,
which look at gene expression levels.  This experimental data can be
stored as a two dimensional data structure, such as a matrix, where rows are
genes and columns are samples.  Linked to this experimental data are meta-data
that give information about the genes studied in the microarray experiment.
For example, this meta-data may be a Gene Ontology graph that gives the
molecular functions of these genes.  These two data sets (the experimental
data and the meta-data) are linked through the genes.  Each gene can have more
than one molecular function and each molecular function can have more than one
gene annotated at that function so there is a many-to-many relationship
between the genes and the molecular functions.  Though these two data sets are
linked, the microarray data and the Gene Ontology data are different data
structures.  The experimental microarray data can be represented as a matrix
and the Gene Ontology meta-data can be represented as a graph.

Now for the original MVC design, where one data set is expected, it is not
clear how to combine these two data sets into one data set.  The meta-data
graph could be stored as a matrix, but a graph is a data structure that can be
well represented as an object with slots for nodes and edges, especially if
the graph is sparse.  As with the relational database tables, it seems
best in terms of data storage, alteration, and retrieval to keep the two data
sets separate.  Also, keeping the two data sets separate does not force a data
structure on the data that may not fit well.

@@

As discussed in Section \ref{Sec:Intro}, the MVC design expects a single
conceptual unit, such as a data frame or a matrix,
with all views referencing this one data set.  However, in Section
\ref{Ssec:Limit} several examples were shown where multiple data sets are
linked.  These examples included linked tables in relational databases and
experimental data linked to meta-data.  Being able to visualize variables
simultaneously from different data sets that are linked would be useful for
exploring relationships between these variables.  Thus, the next step will be
to determine how the MVC design can be revised to account for multiple data
sets.

@@

One possibility for dealing with multiple data sets within the MVC design is to
combine the multiple data sets into one large data set.  With this new
combined data, the original MVC design is appropriate for creating linked
views.  Thus, the strength of combining the data sets into one large data set
is that the MVC paradigm is already appropriately set up to deal with the
one large data set.

Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
multiple data sets into one large data set is not always appropriate or even
possible.  Even if the multiple data sets have the same data structure,
combining the data sets often results in redundant data and makes updating the
data more complicated.  In the situation where the data sets are not of the
same data structure, it may not be possible to combine the data sets into one
data structure that retains all of the original information.  Also, combining
the data into one data structure may give the appearance that some variables
have a relationship that does not exist.  For example, suppose that in a
relational database, table 1's rows are linked to table 2's rows and table 3's
rows are also linked to table 2's rows.  Just because table 1 and table 3 are
both linked to table 2 does not necessarily mean that there is a meaningful
relationship between the data in table 1 and table 3.  However, by combining
the data in these three tables, it would give the appearance that there is a
relationship between the data in table 1 and table 3.  Thus, another
way of linking views based on multiple data sets needs to be designed.

@@

Because of the difficulty of combining multiple data sets, the MVC paradigm
will need to be altered to fit the data.  It has been mentioned that the MVC
design is intended to deal with one data set.  A more straightforward approach
may be to link MVCs.  In linked MVCs, each MVC can be coherent and the
interactions between MVCs can be encapsulated in a set of messages.  One MVC
object to handle each data set.  Thus, within each MVC object, one could
create linked views of that particular data set, which is the same as the
original MVC design.  Then between MVC objects a messaging system would need
to be designed so that a change in one data set would be sent to the other
linked data sets.

@@

\normalsize
As an example, suppose that one has two data sets: one with information at the
state level and one with information at the city level.  These tables are
shown in Table \ref{Tab:CityState}.  The State data include Life Expectancy
(in years from 1970) and Income (per capita from 1974) data
while the City data include Precipitation (in inches from 1975) and Average
High Temperature in January (in Fahrenheit from 2005)
data.  These two data sets are linked by the CityState table that links cities
with the states they reside in.  Now for these two data sets there would be
two MVC objects.  One MVC object for the State data and its views and one MVC
object for the City data and its views.

@@

Two parts of this design will need some special thought: creating an
appropriate messaging system between MVC objects and deciding how the MVC
objects will be related.  In designing an appropriate messaging system between
MVC objects, consideration will be given to what information should pass
between MVC objects and to how messages will propagate between MVC objects.
In deciding the relationship between the MVC objects, consideration will be
given to how the multiple data sets are related.  Clearly, the relationship
between the MVC objects should reflect the relationship between the data
sets.  The two examples of linked data sets that have been discussed
previously were linked tables in a relational database, and experimental data
that are linked to meta-data.  In this paper, the focus will be on experimental
data that are linked to meta-data, as those are the data we are most
interested in visualizing.  In this situation, the experimental data can be
thought of as determining what meta-data are studied.  In the example of
microarray experiment data linked to Gene Ontology meta-data, only Gene
Ontology data that refer to genes in the microarray experiment will be
included in the meta-data.  Thus, there is a hierarchical relationship between
these data sets.  Therefore, the design of the linking between MVC objects
will be based on a hierarchical relationship between the data sets, where one
MVC object creates another MVC object.  These decisions will be discussed in
the following sections as the design for multiple MVC objects is explained.

@@

Implementing an extensible design to create linked, interactive views of
linked data sets will be done in the R language in two packages, called
\Rpackage{MVCClass} and \Rpackage{iSNetwork}.  The \Rpackage{MVCClass} package
will implement the design of the MVC classes and generic functions.  The
\Rpackage{MVCClass} package will define the individual model, view, and
controller classes as well as the \Robject{MVC} class that binds the
individual model, view, and controller classes together into one unit.  The
package will also define the message classes that will allow the components to
communicate with each other.  Thus, the \Rpackage{MVCClass} package will
contain all of the classes necessary for creating linked views.   The goal of
separating these class and generic function definitions into one package is
that they will be available for other R packages that implement linked views
of data, such as the \Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.

With the class definitions created in the \Rpackage{MVCClass} package, the
\Rpackage{iSNetwork} package can decide which classes from \Rpackage{MVCClass}
to use based on the overall goals of the combined software.  The
\Rpackage{iSNetwork} package will also determine how objects of these classes
will behave because the methods for the classes are defined in the
\Rpackage{iSNetwork} package.  In addition to implementing the behavior of
objects, the \Rpackage{iSNetwork} package will also create a graphical user
interface (GUI) so that users can load data, create views, and interact with
the views through menus on the user interface.  As an alternative to the GUI,
users will also have access to the same functionality through command line
functions.

To create the GUI, \Rpackage{iSNetwork} will use the R packages,
\Rpackage{RGtk} and \Rpackage{gtkDevice}.  \Rpackage{RGtk} is an R package
that allows one to interact with Gtk functions from the R interface.  Gtk is
an open-source X Window toolkit \cite{Gtk} for creating user interfaces.  The
\Rpackage{gtkDevice} package creates a Gtk device by turning a Gtk drawing
area into a device that acts like an R device, but still responds to events so
that the view can be interactive.  For more information about these packages,
please see the paper, ``Creating Linked, Interactive Views to Explore
Multivariate Data'' \cite{EW05}.

The next subsections will explain the class definitions and the methods that
pertain to a single model-view-controller object.  Thus, the model classes,
the view classes, the controller, the \Robject{MVC} class (which binds the
model, view and controller classes together), and the message classes that
pass information between the three components in one MVC object will all be
discussed in the following subsections.  All of the class definitions are
implemented in the \Rpackage{MVCClass} package and all of the methods are
defined in the \Rpackage{iSNetwork} package.  Message class definitions that
pertain to communication between the MVC objects will be discussed in Section
\ref{Sec:MultMVC}.

@@

In the MVC design, the model component is the place where data are stored and
updated.  Thus, the model is a storage location that can perform certain
operations.  Information can be sent to the model to add, update, and remove
data and the model can send out messages that changes have occurred to the
data that the model stores.  Thus, the object model for the model classes will
represent both the slots that are needed to store data and the methods that
are needed to perform the operations of the model.

Figure \ref{Fig:Model} shows the inheritance structure for the model classes.
Here the virtual class, \Robject{gModel}, has the slots \Rslot{modelData},
\Rslot{linkData}, \Rslot{virtualData}, and \Rslot{modelName}.  These four
slots are common information that all models will need to store.  The
\Rslot{modelData} slot is the data set for this model, the \Rslot{linkData}
slot is a list of two functions, \Rfunction{toParent} and
\Rfunction{fromParent}, which let this model send information to its parent
model and receive information from its parent model, respectively (see Section
\ref{Ssec:MultLink}), the \Rslot{virtualData} slot is data pertaining to the
views that needs to be stored with the model so that it can be shown in all
views of this model, and the \Rslot{modelName} slot is the name of the model.

@@

As for specific model classes, the classes that are used in the
\Rpackage{iSNetwork} package are \Robject{dfModel}, \Robject{graphModel},
and \Robject{exprModel}.  The \Robject{dfModel} class represents a model where
the \Rslot{modelData} slot has class data frame (or matrix), the
\Robject{graphModel} class represents a model where the \Rslot{modelData} slot
has class graph, and the \Robject{exprModel} class represents a model where the
\Rslot{modelData} slot has class exprSet.

As mentioned previously, the model classes must also perform certain
operations and these operations are reflected in the methods that are defined
for the classes.  All three model classes have an \Rfunction{updateModel}
method that is defined in the \Rpackage{iSNetwork} package.  This method is
called whenever the data need to be updated in response to a user interacting
with a view.  Thus, the \Rfunction{updateModel} method is actually updating the
\Rslot{virtualData} slot because it is this slot in the model classes that
ensures that all views are appropriately visually linked.

Although only three model classes are currently implemented, future users are
welcome to make additions to the model classes for new data structures they
would like to represent.  For example, a user may want to add a new model
class called \Robject{tsModel} to represent time series data.  To make this
addition, a user could extend the \Robject{gModel} class in his or her own
package.

Note that the model classes must also notify views when the data have changed
so that the views can be updated.  The communication between a model and its
views will be discussed in Section \ref{Ssec:OneMess}.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{newViewClass.jpg}}
    \caption{ Inheritance for View Objects. }
    \label{Fig:View}
  \end{center}
\end{figure}
